---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Irina Roybal"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
install.packages("tidyverse")
install.packages("agricolae")

library(tidyverse)
library(agricolae)
library(ggplot2)

getwd()

Lake_ChemPhys_Raw <- 
  read.csv("/home/guest/GITClone/EDE_Fall2025/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",
           stringsAsFactors = TRUE)

Lake_ChemPhys_Raw$sampledate <- 
  as.Date(Lake_ChemPhys_Raw$sampledate, 
          format = "%m/%d/%y")

head(Lake_ChemPhys_Raw$sampledate)
#2
my_theme <-  
  theme(
    plot.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    legend.title = element_text(color = "turquoise4", 
                                face = "bold"),
    legend.text = element_text(color = "turquoise4"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "turquoise4"),
    panel.grid.minor = element_line(color = "turquoise4"),
    plot.title = element_text(face = "bold",
                              color = "turquoise4"),
    axis.title.x  = element_text(
                               face = "italic", 
                               size = 14, 
                               color = "turquoise4"),
    axis.title.y = element_text( 
                               face = "italic", 
                               size = 14, 
                               color = "turquoise4"),
  )


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer: 
H0: Mean lake temperatures recorded during July does not change with depth across all lakes.
Ha: Mean lake temperatures recorded during July change with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
Lake_ChemPhys <- 
  Lake_ChemPhys_Raw%>%
  subset(month(sampledate)==7)%>%
  select(lakename,
         year4,
         daynum,
         depth,
         temperature_C)%>%
  drop_na()
  
#5
library(viridis)

Lake_ChemPhys_Plot <-
  ggplot(Lake_ChemPhys, 
         aes(x = depth,
             y = temperature_C,
             color = temperature_C)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm",
              se = TRUE,
              color = "magenta") +
  xlim(0,20)+
  ylim(0,35)+
  labs(
    title = "Temperature by Depth in July",
    x = "Depth (meters)",
    y = "Temperature (C)")+
   scale_color_viridis(option = "mako", direction = 1) +
  my_theme

print(Lake_ChemPhys_Plot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that in July, as depth increases, temperature decreasees within the lake. The lowest temperatures plateau at around 4 degrees C. So, at a certain point, depth doesn't have any increased impact on the temperature of the water. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
tempdepth.regression <- lm(
  data = Lake_ChemPhys,
  temperature_C ~ depth)

summary(tempdepth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: With an R-squared value of 0.7387, about 73.87% of the variability in temperature is explained by changes in depth. This finding is based on 9726 degrees of freedom which means that there was a large, sufficient sample size to observe this relationship. These results are also statistically significant with a p-value of <2.2e-16, which is far within the threshold of statistical significance: 0.05. With every 1m increase in depth, temperature decreases by around 1.94621 degrees C. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
library(dplyr)
install.packages("corrplot")
library(corrplot)

Lake_ChemPhys_AIC<- lm(data = Lake_ChemPhys_sub,
temperature_C ~ depth + daynum + year4)

step(Lake_ChemPhys_AIC)

#Corrplot for fun

Lake_ChemPhys_sub <-
Lake_ChemPhys %>% 
  select(temperature_C:year4) %>% 
  na.omit()

Lake_ChemPhys_Corr <- cor(Lake_ChemPhys_sub)

corrplot(Lake_ChemPhys_Corr, 
         method = "ellipse")

corrplot.mixed(Lake_ChemPhys_Corr, 
               upper = "ellipse")

#10

Lake_ChemPhys_BestRegression<- lm(data = Lake_ChemPhys_sub,
temperature_C ~ depth + daynum + year4)

summary(Lake_ChemPhys_BestRegression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression are depth, daynum, and year4. This model explains 74.1% of the observed variance with an R-squared value of 0.741. This is a slight improvement over the model using only depth, as that previous model could explain ~0.23% less of the observed variance. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Lake_ChemPhys_ANOVA <- aov(data = Lake_ChemPhys, 
                            temperature_C ~ lakename) 
summary(Lake_ChemPhys_ANOVA)


Lake_ChemPhys_LM <- lm(data = Lake_ChemPhys, 
                       temperature_C ~ lakename)
summary(Lake_ChemPhys_LM)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes. From both the linear model and ANOVA test, the adjusted p-value for the difference in mean temperature among the lakes is statistically significant at <2e-16 (much less than 0.05 or 0.001).



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
Lake_ChemPhys_Plot2 <-
  ggplot(Lake_ChemPhys, 
         aes(x = depth,
             y = temperature_C,
             color = lakename)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm",
              se = TRUE,
              color = "magenta") +
  xlim(0,20)+
  ylim(0,35)+
  labs(
    title = "Temperature by Depth in July of Lakes",
    x = "Depth (meters)",
    y = "Temperature (C)")+
  my_theme

print(Lake_ChemPhys_Plot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lake_ChemPhys_ANOVA)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake and Peter Lake have the same mean temperature, statistically speaking, with a p-value of 0.7827, which is not statistically significant. Paul Lake and Peter Lake also have the same mean temperature with a statistically insignificant p-value of 0.2242. There is no lake that has a mean temperature that is statistically distinct from all other lakes. Each lake has at least one statistically insignificant pairing with another lake (p-value greater than 0.05).


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just looking at Peter and Paul Lake, another test we might explore to see whether they have distinct mean temperatures is a two-sample t-test.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
ChemPhys_Crampton_Ward <- 
  Lake_ChemPhys%>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

ChemPhys_Crampton_Ward

tail(ChemPhys_Crampton_Ward)

Crampton_Ward_ttest <- t.test(ChemPhys_Crampton_Ward$temperature_C
                              ~ ChemPhys_Crampton_Ward$lakename)
Crampton_Ward_ttest
```

>Answer: This test says that the p-value for this relationship is 0.2649. The mean temperature for Crampton Lake is 15.35 degrees C and the mean temperature for Ward Lake is 14.45 degrees C (unequal). This result does not match my answer for part 16. The post-hoc test found that the p-value between Crampton Lake and Ward Lake was 0.9714459. However, both values are statistically insignificant, so in practice, this doesn't change our acceptance of the null hypothesis in terms of the mean temperature of these two lakes. In other words, the mean July temperature of Crampton and Ward Lake is the same.
